Causes and effects in Dichotomous Comparative Judgments: an information-theoretical system of plausible mechanism

### 2024-08-06 _tex/index.tex
Built site for gh-pages

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG), and their associated
structural causal models (SCM), to provide a graphical and formal
representation of the causal structure of a system \citep{Neal_2020}. In
this context, a \emph{graph} denotes a collection of nodes connected by
edges, where nodes represent random variables. The term \emph{directed}
indicates the edges of the graph extend from one node to another, with
arrows showing the direction of causal influence. Moreover, the term
\emph{acyclic} indicates the causal influences do not form a loop,
meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

Regardless of complexity, DAGs can represent various causal structures
using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. The left panels of
Figure~\ref{fig-dags_scms} illustrate these building blocks using the
\emph{standard} representation \citep{Morgan_et_al_2014}. In these
graphs, the variables of interest, or \emph{endogenous} variables
\(V=\{X,Z,Y\}\), are depicted as solid black circles to signify that
they are observed random variables. The arrows in the graphs reflect the
expected direction of causal influences among these variables.

In contrast, the middle panels of Figure~\ref{fig-dags_scms} present the
building blocks using the \emph{magnified} representation. These graphs
incorporate \emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\),
depicted as open circles to signify their unobserved nature. Exogenous
variables represent \emph{errors} or \emph{disturbances} arising from
omitted factors that the investigator chooses not to model explicitly
\citep[27,68]{Pearl_2009}. Thus, they can be viewed as the
`idiosyncratic' causes of each variable. Although these variables are
often omitted for simplicity, as illustrated in the left panels,
including them in the graphs can be beneficial, as their presence can
reveal potential issues related to conditioning and confounding
\citep{Cinelli_et_al_2020}, concepts explored in the next section.

@@

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb3.png}

@@

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb4.png}

@@

@@

\includegraphics[width=0.6\textwidth,height=\textheight]{images/figures/sdag_bb5.png}

@@

@@

\caption{\label{fig-dags_scms}The five fundamental building blocks of
DAGs with corresponding SCMs. \textbf{Note:} left panels show the
standard representation of DAGs, middle panels show the magnified
representation of DAGs, and the right panels show their corresponding
SCMs.}

@@

A careful examination of the left and middle panels of
Figure~\ref{fig-dags_scms} reveals the assumption inherent in these
building blocks. Figures \ref{fig-sdag_bb1} and \ref{fig-mdag_bb1}
depict two unconnected nodes, representing a scenario where variables
\(X\) and \(Y\) are not causally related. Figures \ref{fig-sdag_bb2} and
\ref{fig-mdag_bb2} show two connected nodes, illustrating a scenario
where a parent node \(X\) exerts a causal influence on a child node
\(Y\). Consequently, \(Y\) is considered a \emph{descendant} of \(X\).
Figures \ref{fig-sdag_bb3} and \ref{fig-mdag_bb3} depict a \emph{chain}
(or \emph{pipe}), where \(X\) influences \(Z\), and \(Z\) influences
\(Y\). In this configuration, \(X\) is a parent node of \(Z\), and \(Z\)
is a parent node of \(Y\). Furthermore, the DAG shows that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) completely \emph{mediates} the
relationship between these variables. Figures \ref{fig-sdag_bb4} and
\ref{fig-mdag_bb4} illustrate a \emph{fork}, where variables \(X\) and
\(Y\) are both influenced by \(Z\). In this scenario, \(Z\) is a parent
node of \(X\) and \(Y\). Finally, Figures \ref{fig-sdag_bb5} and
\ref{fig-mdag_bb5} depict a \emph{collider}, also known as
\emph{inmorality}, where variables \(X\) and \(Y\) are concurrent causes
of \(Z\). In this configuration, \(X\) and \(Y\) are not causally
related to each other but both influence \(Z\).

Given the heuristic nature of DAGs, a motivating example can help
illustrate the use of the five fundamental building blocks to construct
a system's causal structure. This example can also help to clarify some
additional conventions for drawing DAGs.

Consider a research problem where the causal effect of a variable \(T\)
on an outcome \(Y\) needs to be investigated. Additionally, the problem
suggests that a variable \(X\) potentially influences both \(T\) and
\(Y\). Such scenarios are not hard to imagine. For instance, \(T\) might
represent different treatments that could affect the recovery from
cancer \(Y\), while \(X\) could denote the patient's age. Similarly, in
the context of a DCJ study like the one described by
\citet{Boonen_et_al_2020}, \(T\) could represent the duration of a
child's cochlear implant use, which might influence the child's overall
speech quality \(Y\), with \(X\) indicating the child's hearing status.
{(not a bad example, but I prefer one using writing skills)}

Figure~\ref{fig-example} presents two graphs illustrating the plausible
causal structure of the motivating example.
Figure~\ref{fig-sdag_example1} shows the \emph{simplified} graph, while
Figure~\ref{fig-mdag_example1} depicts the \emph{magnified} graph of the
DAG. A detailed examination of both figures reveals the presence of at
least four of the five fundamental building blocks.
Figure~\ref{fig-sdag_example1} displays multiple descendants, evident in
pairwise relations such as \(X \rightarrow T\), \(X \rightarrow Y\), and
\(T \rightarrow Y\). It also illustrates a fork with
\(X \rightarrow \{T,Y\}\). Similarly, Figure~\ref{fig-mdag_example1}
features multiple two-unconnected nodes, evident in the pairwise
relations \(\{e_{T} \perp\!\!\!\perp e_{X}\}\),
\(\{e_{T} \perp\!\!\!\perp e_{Y}\}\), and
\(\{e_{X} \perp\!\!\!\perp e_{Y}\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\). The
symbol \(\perp\!\!\!\perp\) denotes a concept know as
\emph{d-separation}, which roughly implies the independence of the
variables. This concept is introduced in the next section.

@@

\includegraphics[width=0.35\textwidth,height=\textheight]{images/figures/sdag_example1.png}

@@

\subcaption{\label{fig-sdag_example1}Simplied DAG}

@@

\includegraphics[width=0.65\textwidth,height=\textheight]{images/figures/mdag_example1.png}

@@

\subcaption{\label{fig-mdag_example1}Magnified DAG}

@@

@@

\subcaption{\label{fig-scm_example1}Collider or inmorality}

>>>>

In statistics, one application incorporating concepts from graph theory
is causal inference. Specifically, the structural approach to causal
inference uses directed acyclic graphs (DAG) to provide a graphical and
formal representation of the causal structure of a system
\citep{Neal_2020}. In this context, a \emph{graph} denotes a collection
of nodes connected by edges, where nodes represent random variables. The
term \emph{directed} indicates the edges of the graph extend from one
node to another, with arrows showing the direction of causal influence.
Moreover, the term \emph{acyclic} indicates the causal influences do not
form a loop, meaning the influences do not cycle back on themselves
\citep{McElreath_2020}.

DAGs offer two key advantages for modeling causality. Firstly, they
represent causal relations in a nonparametric and fully interactive
manner. This generality implies that feasible causal analysis strategies
usually do not require specifying the type of data or the nature of the
functional dependence among variables \citep{Morgan_et_al_2014}.
Secondly, regardless of complexity, DAGs can represent various causal
structures using only five fundamental building blocks
\citep{Neal_2020, McElreath_2020}. These building blocks can be depicted
in three ways: the magnified representation, the standard
representation, and the structural causal model form
\citep[SCM,][]{Morgan_et_al_2014}.

The left panels of Figure~\ref{fig-dags_scms} illustrate the
\emph{magnified} representation. These graphs depict the
\emph{endogenous} variables \(V=\{X,Z,Y\}\) alongside the
\emph{exogenous} variables \(E=\{e_{X},e_{Z},e_{Y}\}\). Endogenous
variables are those whose causal mechanisms the investigator chooses to
model \citep{Neal_2020}. In contrast, exogenous variables represent
\emph{errors} or \emph{disturbances} arising from omitted factors that
the investigator chooses not to model explicitly
\citep[27,68]{Pearl_2009}. The graphs show endogenous variables as solid
black circles to signify that they are observed random variables, while
endogenous variables are depicted as open circles to signify their
unobserved (latent) nature. Lastly, the arrows in the graphs reflect the
expected direction of causal influences among these variables.

Often, the DAGs omit the exogenous variables for simplicity, resulting
in the \emph{standard} representation. However, including exogenous
variables in a graph can be beneficial in some scenarios, as their
presence can reveal potential issues related to conditioning and
confounding \citep{Cinelli_et_al_2020}, concepts explored in the
following section. The standard representation is illustrated in the
middle panels of Figure~\ref{fig-dags_scms}.

@@

Lastly, the right panels of Figure~\ref{fig-dags_scms} depict the SCM
representation of the five fundamental building blocks. SCMs are formal
mathematical models that represent causal relationships within a system
or population \citep{Hitchcock_2024}. An SCM is defined by the set of
endogenous variables \(V\), the set of exogenous variables \(E\), and
the set of functions \(F=\{f_{X},f_{Z},f_{Y}\}\)
\citep{Pearl_2009, Neal_2020}. These functions, referred to as
structural equations, specify each endogenous variable as a
nonparametric function of other variables. Moreover, SCMs use the symbol
`\(:=\)' to indicate the variables' asymmetrical causal dependence and
the symbol `\(\perp\!\!\!\perp\)' to represent \emph{d-separation},
which roughly equates to the concept of variable independence. The
concepts of d-separation and causal (in)dependence are explored in the
following section.

@@

@@

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb3.png}

@@

@@

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb4.png}

@@

@@

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_bb5.png}

@@

\caption{\label{fig-dags_scms}The five fundamental building blocks of
DAGs. \textbf{Note:} left panels show the magnified representation,
middle panels show the standard representation, and the right panels
show their corresponding SCM form.}

@@

A careful examination of Figure~\ref{fig-dags_scms} highlights the
assumptions underlying these building blocks. Figures
\ref{fig-mdag_bb1}, \ref{fig-sdag_bb1}, and SCM \ref{fig-scm_bb1} depict
two unconnected nodes, representing a scenario where variables \(X\) and
\(Y\) are not causally related. Figures \ref{fig-mdag_bb2},
\ref{fig-sdag_bb2}, and SCM \ref{fig-scm_bb2} illustrate two connected
nodes, showing a scenario where a \emph{parent} node \(X\) exerts a
causal influence on a \emph{child} node \(Y\). Consequently, \(Y\) is
considered a \emph{descendant} of \(X\). Figures \ref{fig-mdag_bb3},
\ref{fig-sdag_bb3}, and SCM \ref{fig-scm_bb3} depict a \emph{chain} or
\emph{pipe}, where \(X\) influences \(Z\), and \(Z\) influences \(Y\).
In this configuration, \(X\) is a parent node of \(Z\), and \(Z\) is a
parent node of \(Y\). Furthermore, the DAGs show that \(X\) is an
\emph{ancestor} of \(Y\) and that \(Z\) fully \emph{mediates} the
relationship between these variables. Figures \ref{fig-mdag_bb4},
\ref{fig-sdag_bb4}, and SCM \ref{fig-scm_bb4} illustrate a \emph{fork},
where variables \(X\) and \(Y\) are both influenced by \(Z\). Here,
\(Z\) is a parent node of \(X\) and \(Y\). Finally, Figures
\ref{fig-mdag_bb5}, \ref{fig-sdag_bb5}, SCM \ref{fig-scm_bb5} depict a
\emph{collider}, also known as \emph{inmorality}, where variables \(X\)
and \(Y\) are concurrent causes of \(Z\). In this configuration, \(X\)
and \(Y\) are not causally related to each other but both influence
\(Z\). Additionally, in all SCMs, the errors are assumed to be mutually
independent of each other and from all other variables in the graph, as
evidenced by the pairwise relations \(e_{X} \perp\!\!\!\perp e_{Y}\),
\(e_{X} \perp\!\!\!\perp e_{Z}\), and \(e_{Z} \perp\!\!\!\perp e_{Y}\).

Given the heuristic nature of DAGs, a motivating example can clarify how
to use the five fundamental building blocks to construct the causal
structure of a system. Consider a research problem where the causal
effect of a variable \(T\) on an outcome \(Y\) needs investigation. The
problem suggests that a variable \(X\) potentially influences both \(T\)
and \(Y\). Beyond these relationships, the problem does not specify any
further variables of interest. Such scenarios are not hard to imagine.
For instance, \(T\) might represent different treatments that could
affect the recovery from cancer \(Y\), while \(X\) could denote the
patient's age. Similarly, in the context of a DCJ study like the one
described by \citet{Boonen_et_al_2020}, \(T\) could represent the
duration of a child's cochlear implant use, which might influence the
child's overall speech quality \(Y\), with \(X\) indicating the child's
hearing status. {(not a bad example, but I prefer one using writing
skills)}

Figure~\ref{fig-example} illustrates the plausible causal structure of
the motivating example. A detailed examination of Figures
\ref{fig-mdag_example1}, \ref{fig-sdag_example1}, and SCM
\ref{fig-scm_example1} reveals the presence of at least four of the five
fundamental building blocks. The figures display multiple descendants,
as indicated by pairwise relations such as \(X \rightarrow T\),
\(X \rightarrow Y\), and \(T \rightarrow Y\). Additionally, the figures
features multiple pairs of unconnected nodes, evident from the relations
\(e_{T} \perp\!\!\!\perp e_{X}\), \(e_{T} \perp\!\!\!\perp e_{Y}\), and
\(e_{X} \perp\!\!\!\perp e_{Y}\). Finally, the figures illustrate a fork
with \(X \rightarrow \{T,Y\}\), as well as colliders such as
\(\{X,e_{T}\} \rightarrow T\) and \(\{X,T,e_{Y}\} \rightarrow Y\).

@@

\includegraphics[width=1\textwidth,height=\textheight]{images/figures/mdag_example1.png}

@@

\subcaption{\label{fig-mdag_example1}Magnified representation}

@@

\includegraphics[width=0.55\textwidth,height=\textheight]{images/figures/sdag_example1.png}

@@

\subcaption{\label{fig-sdag_example1}Standard representation}

@@

\[
\begin{aligned}
  X & := f_{X}(e_{X}) \\
  T & := f_{Z}(X,e_{T}) \\
  Y & := f_{Y}(T,X,e_{Y}) \\
  e_{T} & \perp\!\!\!\perp e_{X} \\
  e_{T} & \perp\!\!\!\perp e_{Y} \\
  e_{X} & \perp\!\!\!\perp e_{Y}
\end{aligned}
\]

@@

\subcaption{\label{fig-scm_example1}Structural causal model}
